cardinalR: Generating interesting high-dimensional data structures

### 2025-10-05 paper-cardinalR.Rmd
Removed S-curve and Swiss roll projections (these are common benchmark strutures)

The S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) represents a smooth, non-linear manifold in 3D space. Using `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the geometry is defined by a latent parameter $\theta \sim U(-3\pi/2, 3\pi/2)$, which controls the curvature of the manifold. The first three coordinates form the S-shape:

@@

```{r}
scurve_lang <- langevitour::langevitour(scurve, levelColors = "black", enableControls = FALSE, width = "400px", height = "300px")
```

```{r scurve-proj1}
scaled_data <- scale_data_manual(scurve)

scurve_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r scurve-proj2}

scurve_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.6), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r scurve-proj3}

scurve_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.6, 0.62), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r scurve, eval=knitr::is_html_output(), fig.cap="`langevitour` output  of the `scurve` and `scurvehole` data in $4\\text{-}D$."}

scurvefig <- bscols(
  htmltools::div(
    style = "display: grid; grid-template-columns: 1fr 1fr;",
    scurve_lang
  ),
  device = "xs"
)

class(scurvefig) <- c(class(scurvefig), "htmlwidget")

scurvefig
```

```{r scurve-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `scurve` (a1-a3) and `scurvehole` (b1-b3) data.", fig.width=12, fig.height=4}

scurve_proj1 + scurve_proj2 + scurve_proj3 + 
  plot_layout(ncol = 3, guides = "collect") 
```

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $$X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).$$

@@

```{r}
swissroll_lang <- langevitour::langevitour(swissroll, levelColors = "black", enableControls = FALSE, width = "400px", height = "300px")
```

```{r swissroll-proj1}
scaled_data <- scale_data_manual(swissroll)

swissroll_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll-proj2}

swissroll_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.65), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll-proj3}

proj_obj3 <- get_projection(projection = projection, 
                            proj_scale = 1.2, 
                            scaled_data = scaled_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 1.5, 
                                              axis_pos_x = -0.55, 
                                              axis_pos_y = -0.55, 
                                              threshold = 0.016))

swissroll_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.5), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll, eval=knitr::is_html_output(), fig.cap="`langevitour` output  of the `swissroll` data in $3\\text{-}D$."}

swissrollfig <- bscols(
  htmltools::div(
    style = "display: grid; grid-template-columns: 1fr 1fr;",
    swissroll_lang
  ),
  device = "xs"
)

class(swissrollfig) <- c(class(swissrollfig), "htmlwidget")

swissrollfig
```

```{r swissroll-proj, eval=knitr::is_latex_output(), fig.pos="H", fig.cap="Three $2\\text{-}D$ projections from $3\\text{-}D$, for the `swissroll` data."}
#| fig-width: 12
#| fig-height: 4

swissroll_proj1 + swissroll_proj2 + swissroll_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

>>>>

The S-curve structure represents a smooth, non-linear manifold in 3D space. Using `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the geometry is defined by a latent parameter $\theta \sim U(-3\pi/2, 3\pi/2)$, which controls the curvature of the manifold. The first three coordinates form the S-shape:

@@

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension. The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $$X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).$$

@@

### 2025-10-05 paper-cardinalR.tex
Removed S-curve and Swiss roll projections (these are common benchmark strutures)

The S-curve structure (Figure \ref{fig:scurve-proj}) represents a smooth, non-linear manifold in 3D space. Using \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the geometry is defined by a latent parameter \(\theta \sim U(-3\pi/2, 3\pi/2)\), which controls the curvature of the manifold. The first three coordinates form the S-shape:

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/scurve-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `scurve` (a1-a3) and `scurvehole` (b1-b3) data.}\label{fig:scurve-proj}
\end{figure}

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \[X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).\]

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/swissroll-proj-1} \caption{Three $2\text{-}D$ projections from $3\text{-}D$, for the `swissroll` data.}\label{fig:swissroll-proj}
\end{figure}

>>>>

The S-curve structure represents a smooth, non-linear manifold in 3D space. Using \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the geometry is defined by a latent parameter \(\theta \sim U(-3\pi/2, 3\pi/2)\), which controls the curvature of the manifold. The first three coordinates form the S-shape:

@@

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension. The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \[X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).\]

@@
